Studying public transit API query logs\\ to get an indication of travel flows

### 2016-01-08 paper.tex
Modified related work

Visualizations are crucial to understanding the patterns in data representing transit flows.
Origin Destination (OD) type representations are used when the origin and destination preside over intermediate locations on the path of movement. 
The OD Matrix is most common visualization technique in this type of representation. 
Rows and columns correspond to locations while cells are coloured to express the volume of flow.
OD matrices can be reordered to emphasise connectivity between locations.
While this feature is particularly useful for identifying trends and outliers in the data, the tabular display is challenging for lay users to interpret and less expressive of connections between locations than representations based on the node-link metaphor.
In recent years, such representations have become increasingly prominent for displaying movement in scientific publications as well as popular culture.
Locations are depicted as nodes and arrows are drawn between them to indicate movement in a certain direction, allowing for exact and individual relationships to be traced.
\todo{PC: Alvin, can we have a citation for this paragraph? Maybe you can cite one your own papers here?}

@@

%\todo{Maybe a sub header here?} - PC: I wouldn't do a subheader...
We are particularly interested in studying the potential application of \emph{route planning} queries as an information resource for flow analysis due to the input it captures from human users.
The other features are less relevant for this purpose as they are generally polled by digital signage providers or status boards, and thus do not reflect direct human demands.
The dataset described in this paper is compiled from the \emph{Apache} \emph{access.log} files generated between October 2012 and December 2015, filtered by ``connections'' entries.
We made a copy of this dataset available in CSV format can be downloaded at \todo{\url{http://datawijs.be/apilog.tar.gz}}.
%CSV header of the final logs: datetime,requested_datetime,from_id,from_name,from_long,from_lat,to_id,to_name,to_long,to_lat,user_agent,operating_system,is_bot
Each row within this dataset contains:
\begin{enumerate}
\item A timestamp of when the request was executed,
\item The request path between two stations and their associated geographic coordinates,
\item The user-agent string cfr. RFC2616,
\item The operating system of the user-agent and
\item A flag for whether the user-agent is a bot.
\end{enumerate}

@@

Flow analysis involves the comparison of aggregated movement between distinct locations within a specific time frame. 
In most cases, the analytical objective is to identify trends that occur in journeys made between pairs of locations and allow for the sensitivities to be explored. 
Data visualization is typically employed to facilitate in this process thus the input data must be transformed into a format that allows for the quantity and directional flow of movement to be visually encoded.
Several stages of data processing are required to arrive at this outcome.

@@

As we were specifically interested in studying morning and evening rush hour travel that occur on weekdays, we filtered the dataset to exclude data records created outside the time range of 06:00 hours and 10:00 hours as well as 17:00 hours and 21:00 hours on weekdays, and all data records created on weekend.
Queries made by automated user-agents like search engine bots and data harvesters can be considered as valid traffic but may over represent connections between certain stations.
Such queries have been excluded since we are specifically interested in studying human travel behavior.
Station names tend to contain be highly inconsistent since any text information can be parsed to the API.
This may include invalid character encoding, spelling mistakes and the use of unstandardized abbreviations.
We employ an open-source reconciliator (\url{https://github.com/irail/stations}) to reduce spelling variations and remove data records with missing information. 
Additionally, it automatically links station names to their corresponding geographical coordinates.

@@

Next, the data is aggregated so that province level flow patterns are emphasized.
\cref{fig:method} provides a diagrammatic representation of spatial aggregation on the basis of provincial administrative boundaries.
To compare the volume of flow between rural provincial areas and urbanized zones, train stations in major cities are aggregated seperately. 
If an origin or destination station is a located outside of Belgium, it is considered international travel and aggregated in a distinct group. 
A diagrammatic breakdown is depicted in \cref{fig:method-c} where color and shape is used to demarcate each group.
The following types of flows are observable from the outcome of aggregation:

@@

\subsection{Visualization}

Visualization is crucial for making data analysis tangible, so that results can be communicated and debated \cite{robinson2008collaborative}. 
This is key for our research, since the questions presented are exploratory in nature \cite{kraak2008exploratory} and can be addressed in many ways. 
Movement data has received substantial attention from the visualization community and a range of techniques has been developed to support analysis \cite{andrienko2012visual}. 
Of these techniques, flow visualizations are commonly employed to facilitate the comparison of aggregated movement. 
Flow visualizations can be broadly categorized into two groups. Representations that preserve the complete trajectory and Origin Destination type representations that take only the start and end locations into account.
Since our data solely consists of OD information, we narrow our description to the latter group of visualization techniques.

The chord diagram is a visualization technique based on the node-link metaphor that arranges nodes along the circumference of a circle. 
Each node is represented by an arc where its length is proportional to the total volume of incoming and outgoing flows. 
Chords or curved line segments are drawn to connect nodes. 
The width at the head or tail of each chord indicates the amount of movement relative frequency of movement from a certain location to another.
In our implementation of the chord diagram, nodes are colour coded to indicate individual provinces and the major cities located within their administrative boundaries. 
Interactive filtering is introduced to simplify the visualization and details are provided in pop-up dialogue when upon mouse-over.

@@

The chord diagrams shown in \cref{fig:brussels} and \cref{fig:antwerp} depict the aggregated number of queries made between any two stations in the morning and evening respectively. 
Visualizing our dataset in this manner reveals the complexity of flows on the Belgian rail system and the significance of cities in weekday travel.
Turning our attention to morning commutes, we can observe that Antwerp, Brussels and Gent are the dominant destinations.
The reverse occurs in the evening when trips are made towards provincial stations.
Brussels stands out as the most distinctive visual element in both figures, indicating its function as the principal centre of rail activity.
On a regional level, its function as a centre appears to be more distinctive for the Walloon region than Flanders.
With exception to Liege, queries from the Walloon region are generally made from provincial stations towards Brussels instead of major cities within their administrative boundaries. 
Queries from Flanders, on the hand, tend to be distributed among four major cities otherwise known as the Flemish Diamond, a network of metropolitan areas in Belgium comprised of Ghent, Brussels, Antwerp and Leuven. 
These insights indicate that the structure of flows in Flanders appear to follow a polycentric pattern while that of the Walloon region is relatively monocentric, with   Brussels as a major centre.
The difference between both regions appear to correspond with existing measures of population density and degree of urbanisation, providing valuable insight as well as alternative perspectives into the function of cities in rural and urban settings.

>>>>

%Visualizations are crucial for understanding the patterns in data and representing transit flows.
Visualizations are crucial for making data analysis tangible, so that results can be communicated and debated \cite{robinson2008collaborative}. 
This is key for our research, since the questions presented are exploratory in nature \cite{kraak2008exploratory} and can be addressed in many ways. 
Movement data has received substantial attention from the visualization community and a range of techniques has been developed to support flow analysis \cite{andrienko2012visual}. 
Flow visualizations can be broadly categorized into two groups. 
Representations that display the complete trajectory and Origin Destination (OD) type representations that take only the start and end locations into account.
Since our data solely consists of OD information, we narrow our description to the latter group of visualization techniques.
OD type representations are commonly employed when the origin and destination preside over intermediate locations on the path of movement.

The OD Matrix is most common visualization technique based on this type of representation. 
Rows and columns correspond to locations while cells are coloured to express the volume of flow.
OD matrices can be reordered to emphasise connectivity between locations.
While this feature is particularly useful for identifying trends and outliers in the data, the tabular display is challenging for lay users to interpret and less expressive of connections between locations than representations based on the node-link metaphor.
In recent years, such representations have become increasingly prominent for displaying movement in science as well as popular culture.
Locations are depicted as nodes and arrows are drawn between them to indicate movement in a certain direction, allowing for exact and individual relationships to be traced. 
This representation is detailed and straightforward to understand but becomes cluttered when multiple lines intersect. 
A number of solutions have been proposed for clutter reduction. 
Filtering is the most commonly practiced. 
The same goal can be achieved by clustering lines that have similar properties like common start, end or intermediate locations \todo{Andrienko, 2007 -198}. 
Another solution is to transform the visual representation. 
Flow maps \todo{Phan, 2005 -88} reduces clutter by rerouting the arrows to minimize intersection and obtain layouts that emphasize key geographic features such as roads, rivers or coastlines. 
\todo{Alvin: Will expand abit more to describe why chord diagrams were chosen}

@@

We are particularly interested in studying the potential application of \emph{route planning} queries as an information resource for flow analysis due to the input it captures from human users.
The other features are less relevant for this purpose as they are generally polled by digital signage providers or status boards, and thus do not reflect direct human demands.
The dataset described in this paper is compiled from the \emph{Apache} \emph{access.log} files generated between October 2012 and December 2015, filtered by ``connections'' entries.
We made a copy of this dataset available in CSV format can be downloaded at \todo{\url{http://datawijs.be/apilog.tar.gz}}.
%CSV header of the final logs: datetime,requested_datetime,from_id,from_name,from_long,from_lat,to_id,to_name,to_long,to_lat,user_agent,operating_system,is_bot
Each row within this dataset contains:
\begin{enumerate}
\item A timestamp of when the request was executed,
\item The request path between two stations and their associated geographic coordinates,
\item The user-agent string cfr. RFC2616,
\item The operating system of the user-agent and
\item A flag stating if the user-agent is a bot.
\end{enumerate}

@@

Flow analysis involves the comparison of aggregated movement between distinct locations within a specific time frame. 
In most cases, the analytical objective is to identify trends that occur in journeys made between pairs of locations and allow for the sensitivities to be explored. 
Data visualization is typically employed to facilitate in this process thus the input data must be transformed into a format that allows for the quantity and directional flow of movement to be visually encoded.
The data must undergo several stages of data processing to arrive at this outcome.

@@

We were specifically interested in understanding how travel during the morning rush hour on weekdays differ from weekends. In this regard, the dataset was filtered to exclude data records created beyond the time range of 06:00 hours and 10:00 hours.
Queries made by automated user-agents like search engine bots and data harvesters can be considered as valid traffic but may over represent connections between certain stations.
Accordingly, such queries were excluded since they do not explicitly represent human travel behavior.
Station names tend to contain be highly inconsistent since any text information can be parsed to the API.
This may include invalid character encoding, spelling mistakes and the use of unstandardized abbreviations.
We employed an open-source reconciliator (\url{https://github.com/irail/stations}) to reduce spelling variations and remove data records with missing information. 
Through this process, station names were also linked to their corresponding geographical coordinates.

@@

Next, the data is aggregated so that province level flow patterns are emphasized.
\cref{fig:method} provides a diagrammatic representation of spatial aggregation on the basis of provincial administrative boundaries.
To compare the volume of flow between rural provincial areas and urbanized zones, train stations in major cities were aggregated into seperate groups. 
Simiarly, stations located outside of Belgium were aggregated to distinct group. 
A diagrammatic breakdown is depicted in \cref{fig:method-c} where color and shape is used to demarcate each group.
The following types of flows are observable from the outcome of aggregation:

@@

\subsection{Visualization}
%Flow visualizations are commonly employed to facilitate the comparison of aggregated movement. 
The chord diagram is a visualization technique based on the node-link metaphor that arranges nodes along the circumference of a circle. 
Each node is represented by an arc where its length is proportional to the total volume of incoming and outgoing flows. 
Chords or curved line segments are drawn to connect nodes. 
The width at the head or tail of each chord indicates the amount of movement relative frequency of movement from a certain location to another.
In our implementation of the chord diagram, nodes are colour coded to indicate individual provinces and the major cities located within their administrative boundaries. 
Interactive filtering is introduced to simplify the visualization and details are provided in pop-up dialogue when upon mouse-over.

@@

The chord diagrams shown in \cref{fig:mornings} depict the aggregated number of queries made between any two stations on weekdays (See \cref{fig:weekdays_morning}) and weekends (See \cref{fig:weekends_morning}) respectively. 
Visualizing our dataset in this manner reveals the complexity of flows on the Belgian rail system and the significance of cities in daily travel.
%We can observe that Brussels, Antwerp, Gent and Leuven are the dominant destinations.
%Turning our attention to weekday travel, we can observe that Antwerp, Brussels and Gent are the dominant destinations.
%The reverse occurs in the evening when trips are made towards provincial stations.
Brussels City stands out as the most distinctive visual element in both figures, indicating its function as the principal centre of rail activity (see \cref{fig:mornings}a).
On a regional level, its function as a centre appears to be more distinctive for the Walloon region than Flanders.
With exception to Liège city (see \cref{fig:mornings}e), queries from the Walloon region are generally made from provincial stations towards Brussels instead of major cities within their administrative boundaries. 
Queries from Flanders, on the hand, tend to be distributed among four major cities otherwise known as the Flemish Diamond, a network of metropolitan areas in Belgium comprised of Brussels, Antwerp (see \cref{fig:mornings}b), Gent (see \cref{fig:mornings}c) and Leuven (see \cref{fig:mornings}d). 
These insights indicate that the structure of flows in Flanders appear to follow a polycentric pattern while that of the Walloon region is relatively monocentric, with Brussels as a major centre.
The difference between both regions appear to correspond with existing measures of population density and degree of urbanisation, providing valuable insight as well as alternative perspectives into the function of cities in rural and urban settings.
